---
title: "hmdbQuery: working with Human Metabolome Database (hmdb.ca)"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{hmdbQuery: working with Human Metabolome Database (hmdb.ca)}
output:
  html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
---

# Initial remarks

The human metabolomics database (HMDB, http://www.hmdb.ca)
includes XML documents describing 114000 metabolites.  We will show
how to manipulate the metadata on metabolites fairly
flexibly.

```{r setu,echo=FALSE,results="hide"}
suppressMessages({
suppressPackageStartupMessages({
library(hmdbQuery)
library(gwascat)
})
})
```

# Key utilities of the package

The hmdbQuery package includes a function for querying HMDB
directly over HTTP:

```{r lk1}
library(hmdbQuery)
lk1 = HmdbEntry(prefix = "http://www.hmdb.ca/metabolites/", 
       id = "HMDB0000001")
```
The result is parsed and encapsulated in an S4 object
```{r lk2}
lk1
```

The size of the complete import of information about a single metabolite
suggests that it would not be too convenient to have comprehensive
information about all HMDB constituents in memory.  The most effective
approach to managing the metadata will depend upon use cases to be
developed over the long run.

Note however that this package does provide snapshots
of certain direct associations derived from all available
information as of Sept. 23 2017.  Information
about direct associations reported in the database
is present in tables `hmdb_disease`,
`hmdb_gene`, `hmdb_protein`, `hmdb_omim`.  For
example
```{r lkta}
data(hmdb_disease)
hmdb_disease
```

# Working with the metadata

## Disease associations

Some HMDB metabolites have been mapped to diseases. 

```{r lkdis3}
d1 = diseases(lk1) # matrix
pmids = unlist(d1[3,1][[1]]["pubmed_id",])
library(annotate)
pm = pubmed(pmids[1])
ab = buildPubMedAbst(xmlRoot(pm)[[1]])
ab
```

## Biospecimen and tissue location metadata

Note that pre HMDB v 4.0, biospecimens were called biofluids.

There are arbitrarily many biospecimen and tissue associations
provided for each HMDB entry.  We have direct accessors,
and by default we capture all metadata, available through
the `store` method.

```{r lkdee}
biospecimens(lk1)
tissues(lk1)
st = store(lk1)
head(names(st))
length(names(st))
st$protein_assoc["name",]
st$protein_assoc["gene_name",]
```

